---
title: "Brown Datathon 2018 - RI Electricity"
author: "Thoa Ta"
date: "March 3, 2018"
output: html_notebook
---

```{r setup}
# include the necessary libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(scales)

# set all the plots in this document to white background
theme_set(theme_bw())
```

```{r set-path, include=FALSE}
path <- "C:/Users/Thoa Ta/Downloads/brown_datathon_2018/"
setwd(path)
```

```{r staging}
# load & check data
mydata <- read_csv("my_copy_of_2017_smd_hourly.csv")
mydata <- mydata %>% select(-13,-14)
head(mydata)
summary(mydata)

# check missing data: all zero sums means no missing data
colSums(is.na(mydata))

# check correlation
cor(mydata[,3:12])
```

We see that the following pairs have the highest correlations:

Variable A | Variable B | Correlation
--- | --- | ---
RT_Demand | DA_Demand | .97
DA_EC | DA_LMP | .99
RT_EC | RT_LMP | .99

Since I don't have background in the electricity market, variables other than the Real-Time Demand and the Day-Ahead Demand make little sense to me. Hence, I chose to work on the Demand variables only.

## Electricity demand by month

```{r group-by-month}
# parse the Date column
mydata$Date <- dmy(mydata$Date)
head(mydata)

mydata %>% 
  mutate(Month = as.factor(month(Date))) %>%
  group_by(Month) %>%
  summarize("Real-Time Demand" = mean(RT_Demand),
            "Day-Ahead Demand" = mean(DA_Demand)) %>%
  gather(2:3, key = Type, value = Mean_Demand) %>%
  ggplot(aes(x = Month, y = Mean_Demand, color = Type)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("green", "blue")) +
  labs(title = "Average Electricity Demand by Month in 2017", y = "Average Demand")
```
This plot shows us two things:
    
  1. How close the average day-ahead and real-time demand are in some months, and
  2. The demands in different months throughout a year.

We will further examine these two observations below. 
  
### 1. Variance in demand and day-ahead / real-time correlation

I wonder if the variance in monthly demand could be indicative of the day-ahead / real-time correlation. In other words, can we say that day-ahead and real-time values are less correlated in months with higer variance?

```{r month-variance}
mydata %>%
  mutate(Month = as.factor(month(Date))) %>%
  group_by(Month) %>%
  summarize("Real-Time Demand" = mean(RT_Demand),
            "Day-Ahead Demand" = mean(DA_Demand),
            "Real-Time Std.Err" = sd(RT_Demand) / sqrt(n()),
            "Day-Ahead Std.Err" = sd(DA_Demand) / sqrt(n())) %>%
  gather(2:3, key = Demand_Type, value = Mean_Value) %>%
  gather(2:3, key = Std.Err_Type, value = Std.Err_Value) %>%
  arrange(Month) %>%
  filter((Demand_Type == "Real-Time Demand" & Std.Err_Type == "Real-Time Std.Err") |
           (Demand_Type == "Day-Ahead Demand" & Std.Err_Type == "Day-Ahead Std.Err")) %>%
  ggplot(aes(x = Month, y = Mean_Value, color = Demand_Type)) +
  geom_pointrange(aes(ymin = Mean_Value - Std.Err_Value, ymax = Mean_Value + Std.Err_Value)) +
  scale_color_manual(values = c("green", "blue")) +
  labs(title = "Average Electricity Demand by Month in 2017", 
       subtitle = "With standard errors", 
       y = "Average Demand", 
       color = "Type")
```

It seems like the relationship is not that strong.

### 2. Demand trend throughout the year

We see that electricity demand is the highest in July and August; relatively high in December, January, February, March, June, and September; and lowest in April, May, October, and November. Possible explanations are as follows:

- July and August are the peak of summer, so electricity for running the air-conditioner is high.

- April-May and October-November are the transition months from a cold season to a hot one (and vice versa), so users might simply turn off the heater and not (yet to / need to) turn on the air-conditioner. Therefore, electricity demand is low.

- June and September are the pre and post of summer time, so electricity demand is slightly, but not drastically, higher than usual.

- The other months show the average electricity need for heaters during winter time. 

**One interesting point for future research** would be: consider what types of appliances are used by the majority of users *(probably both industrial and residential users?)* in the summer and the winter seasons, to verify the demand for electricity.

## Electricity demand by time of day

Next up, I want to examine the demand trend throughout the course of a day, faceted by month.

*Note* Since there is not a lot of difference between the Real-Time and Day-Ahead Demand, I will just use the Real-Time Demand for the following analysis.

```{r, group-by-time-month-together}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  #filter(Month %in% c("12", "1", "2", "3")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(title = "Average Electricity Demand in 2017", 
       subtitle = "Months together", 
       y = "Average Real-Time Demand")
```

To see the trend in each month more clearly, I am going to facet the data into month windows.

```{r, group-by-time-month-separated}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  #filter(Month %in% c("12", "1", "2", "3")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  facet_wrap( ~ Month) +
  theme(legend.position = 'none') +
  labs(title = "Average Electricity Demand in 2017", 
       subtitle = "12 windows represent 12 months",
       y = "Average Real-Time Demand")
```

From the two plots above, we observe that :

  1. The timewise trends group themselves into five (5) groups of pattern:
      a. December to March
      b. April and May
      c. June and September
      d. July and August
      e. October and November
  
  2. All 5 groups have similar lowest points of demand (the 3-5am range), but their peak patterns differ. 
  
Below, we will look more closely at each group to examine their peak pattern.

### December to March
```{r dec-to-march}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  filter(Month %in% c("12", "1", "2", "3")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  scale_color_manual(values = c("#F8766D", "#DE8C00", "#B79F00", "#FF65B0")) +
  scale_fill_manual(values = c("#F8766D", "#DE8C00", "#B79F00", "#FF65B0")) +
  labs(title = "Average Electricity Demand in 2017 Months: 1,2,3,12",
       y = "Average Real-Time Demand")
```

### April and May
```{r apr-and-may}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  filter(Month %in% c("4", "5")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  scale_color_manual(values = c("#7CAE00", "#00BA38")) +
  scale_fill_manual(values = c("#7CAE00", "#00BA38")) +
  labs(title = "Average Electricity Demand in 2017 Months: 4,5",
       y = "Average Real-Time Demand")
```

### June and September
```{r jun-and-sep}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  filter(Month %in% c("6", "9")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  scale_color_manual(values = c("#00C08B", "#619CFF")) +
  scale_fill_manual(values = c("#00C08B", "#619CFF")) +
  labs(title = "Average Electricity Demand in 2017 Months: 6,9",
       y = "Average Real-Time Demand")
```

### July and August
```{r jul-and-aug}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  filter(Month %in% c("7", "8")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  scale_color_manual(values = c("#00BFC4", "#00B4F0")) +
  scale_fill_manual(values = c("#00BFC4", "#00B4F0")) +
  labs(title = "Average Electricity Demand in 2017 Months: 7,8",
       y = "Average Real-Time Demand")
```

### October and Novemeber
```{r oct-and-nov}
mydata %>% 
  mutate(Month = as.factor(month(Date)), Hour = as.numeric(Hr_End)) %>%
  select(-Hr_End) %>%
  filter(Month %in% c("10", "11")) %>%
  group_by(Month, Hour) %>%
  summarize(Mean_RT_Demand = mean(RT_Demand),
            Se_RT_Demand = sd(RT_Demand) / sqrt(n())) %>%
  ggplot(aes(x = Hour)) +
  geom_line(aes(y = Mean_RT_Demand, color = Month)) +
  geom_ribbon(aes(ymin = Mean_RT_Demand - Se_RT_Demand, 
                  ymax = Mean_RT_Demand + Se_RT_Demand, 
                  fill = Month), 
              alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  scale_color_manual(values = c("#C77CFF", "#F564E3")) +
  scale_fill_manual(values = c("#C77CFF", "#F564E3")) +
  labs(title = "Average Electricity Demand in 2017 Months: 10,11",       
       y = "Average Real-Time Demand")
```


## Next steps:

1. **Shift the color wheel** to make blue shades align with cold months and red shades align with hot months.
